\tcode{Generalisation of nth_element to a range of nths}

### 2021-12-28 draft_lundberg_nth_elements.tex
rebase wording to new standards draft, cleanup wording and...
use upper_bound in example implementation, as in ref impl repo

\begin{document}
\vspace*{-9em}
\title{\tcode{Generalisation of nth_element to a range of nths}}
\author{
Johan Lundberg\\
}
\date{} %unused. Type date explicitly below.
\maketitle
\vspace*{-5em}

\begin{tabular}{ll}
Document \#:& D2375R1\\
Date: &2021-12-26 \\
Project: & Programming Language C++ \\
Audience: & LEWG \\
Email: &\href{mailto:lundberj@gmail.com}{lundberj@gmail.com}
\end{tabular}

@@

\subsection*{Revision history vs R0}

2021-12-16: Added clarifications and more background, explanations, wording clarifications, references, summarized performance study, applications, more questions/answers. --Based on feedback and material sent to SG9 list 2021-12-09. Synopsis: \tcode{S last} argument was missing 2021-12-19.
2021-12-25: Simplified wording (change marked at section \ref{synopsis}). Added images for equalization example. Formatting and typos. Added example of \emph{Before} with projections, and example \emph{Before} with fixed size m=5.\newline 
2021-12-26: Added explanatory images to the introduction. More thoughts on naming and minor text clarifications.

@@

The use and analysis of such algorithms is widespread and mature\cite{Alsuwaiyel2001,Panh2002,lent1996,Shen1997} and is available to Python programmers as \tcode{numpy.partition}\cite{NpPart,NPImpl} since 2014.

@@

A possible implementation of the range-of-nths algorithm is provided (section \ref{Implementation}).
It translates naturally to \tcode{std::ranges} versions just like \tcode{std::nth_element}, returning \tcode{last}.

@@

To clarify what is new, the addition is here called \tcode{std::multi_nth_element}, but is proposed to go as new overloads for \tcode{std::nth_element}.

@@

As example data, consider the permuted integers [100,126) at index [0,26) and the sorted counterpart:\\[0.41em]
\includegraphics[width=0.999\textwidth]{plotting/figs/rnd.png}\\[0.41em]
\includegraphics[width=0.999\textwidth]{plotting/figs/sort.png}\\[0.41em]

@@

What the current-standard \tcode{std::nth_element} does is to arrange and partition the data as described in the previous section. For example, with \tcode{nth = begin+7}, the element in the position pointed to by nth is the element that  would  be  in  that position if the whole range were sorted, and all subsequent values are no less than that value:

@@

This proposal adds the possibility to provide a range of nths, such as \tcode{\{begin+7,begin+20\}} :

@@

or at \tcode{\{7,12,20\}}:

@@

or  at \tcode{\{5,6,14,15\}}:

@@

The proposal can be implemented with an algorithm that partitions the data on the midpoint of nths using the single-nth \tcode{nth_element} and proceeds to recurse into the two partitions. Ref \cite{Alsuwaiyel2001} considers this algorithm and concludes
 complexity \mbox{O(N log m)} where \mbox{N = last - first} and m is the number of unique elements in nths.
The same paper presents a parallel version, building on refs \cite{Akl1984,Akl1989,Shen1997}.

@@

A possible implementation\footnote{
The implementation is not the point of the proposal but it may help explain it. Feedback would be very much appreciated. 
The algorithm described in \cite{Alsuwaiyel2001} is a little bit simpler since it takes \emph{unique} nths, something that is not reasonable to require here. The algorithms is \mbox{O(N log m)} except for a small N-independent term \mbox{O(M-m)} where M is the size of nths, due to the iterator comparisons and increments in \tcode{find_if_not} to skip over doublets in nths. The first if-statement is found in existing single-nth \tcode{nth_element} implementations, but here it also prevents some unnecessary bisections of nths.}\cite{p2375RefImpl}
is shown below. Comparisons and projections can also be fed through in the most natural way.

\begin{codeblock}
template< class RandomAccessIterator, class RandomAccessIterator2 >
constexpr void @\textbfx{multi_nth_element}@(RandomAccessIterator first, 
  RandomAccessIterator2 nth_first, RandomAccessIterator2 nth_last, 
  RandomAccessIterator last)
{
  if (last - first <= 32) { @\textbfx{std::sort(first, last)}@; return; }
  const auto nth_dist = nth_last - nth_first;
  if (nth_dist == 0 || *nth_first == last) return;
  const auto nth_mid = nth_first + nth_dist / 2;
  const auto at_nth_mid = *nth_mid;
  @\textbfx{nth_element(first, at_nth_mid, last)}@;
  @\textbfx{multi_nth_element(first, nth_first, nth_mid, at_nth_mid)}@;
  if (at_nth_mid != last){
    const auto nth_left = std::find_if_not(nth_mid, nth_last, 
      [at_nth_mid](auto v) {return v == at_nth_mid; });
    @\textbfx{multi_nth_element(at_nth_mid + 1, nth_left, nth_last, last)}@;
  }
}

@@

\hspace{3ex}\begin{tabular}{|l|} 
  \hline 
  \textbf{Before} \\
  \hline 
\begin{codeblock} 
nth_element(v.begin(), nths[5/2-1], v.end(), pred);
nth_element(v.begin(), nths[(5/2)/2-1], nths[5/2-1], pred);
nth_element(nths[(5/2)/2-1]+1, nths[5/2-1], v.end(), pred);
nth_element(nths[(5/2)-1]+1, nths[(5/2)/2+5/2-1], nths[5-1]), pred);
nth_element(nths[5-1]+1, nths[(5/2)/2+5/2-1], v.end(), pred);
\end{codeblock} 
\\
\hline 
\end{tabular}

@@

\begin{codeblock}
std::nth_element({a,b}, v);
processPage(a,b); // May also now sort the small subrange with std::sort(a,b);
\end{codeblock}

@@

\newpage
\section{Wording and Synopsis }

\subsection{First wording draft \protect{\tcode{[alg.nth.element]}}}

\added{Underlined green text} marks additions.

\label{wording}\label{synopsis}

\textbf{1} Let comp be less{} and proj be identity{} for the overloads with no parameters by those names.

\textbf{2} Preconditions: [first, nth) and [nth, last) are valid ranges. For the overloads in namespace std, RandomAccessIterator meets the Cpp17ValueSwappable requirements ([swappable.requirements]), and the type of *first meets the Cpp17MoveConstructible (Table 30) and Cpp17MoveAssignable (Table 32) requirements.
\added{For the overloads taking a range [nths_first,nths_last), \newline RandomAccessIterator2 is a RandomAccess iterator, and *nths_first is convertible to \newline RandomAccessIterator.}

\textbf{3} Preconditions: The elements e of [first, last) are partitioned with respect to the expression bool(invoke(comp, invoke(proj, e), value)).

\textbf{4} Effects: After nth_­element the element in the position pointed to by nth is the element that would be in that position if the whole range were sorted with respect to comp and proj, unless nth==last. Also for every iterator i in the range [first, nth) and every iterator j in the range [nth, last) it holds that: bool(invoke(comp, invoke(proj, *j), invoke(proj, *i))) is false. 
\added{For the overloads taking a range of nths, this holds for all nth in nths}$^\ddagger$

\textbf{5} 
Complexity: For the overloads with no ExecutionPolicy, linear on average.
For the overloads with an ExecutionPolicy, O(N) applications of the predicate, and O(N log N) swaps, where N = last - first. 
\added{For overloads taking a range [nths_first,nths_last) but no ExecutionPolicy the complexity is approximately \mbox{O(N log m + M)} where m is the number of unique elements in [nths_first,nths_last), and M = nths_last - nths_first.
For overloads taking a range [nths_first,nths_last) and an ExecutionPolicy the complexity is \mbox{O(N log N + M)}.}

$\ddagger$: \textbf{Note:} Rev R0 had an different wording with the same intention, but was redundant, as well as incomplete with respect to \emph{comp} and \emph{proj}.

@@

Added signatures \tcode{std::}

@@

template<class RandomAccessIterator, class RandomAccessIterator2, class Compare>
constexpr void nth_element(
RandomAccessIterator first, 
RandomAccessIterator2 nths_first, RandomAccessIterator2 nths_last,
RandomAccessIterator last, Compare comp);

template<class ExecutionPolicy, class RandomAccessIterator, class RandomAccessIterator2>
void nth_element(ExecutionPolicy&& exec,
RandomAccessIterator first, 
RandomAccessIterator2 nths_first, RandomAccessIterator2 nths_last,
RandomAccessIterator last);

@@

Added signatures \tcode{std::ranges::}

\begin{codeblockAdd}
namespace ranges {
  template<random_access_iterator I, sentinel_for<I> S, 
  random_access_range R2, class Comp = ranges::less, class Proj = identity>
  requires sortable<I, Comp, Proj>
  constexpr I nth_element(I first, R2&& nths, S last, Comp comp = {}, Proj proj = {});

  template<random_access_range R, 
  random_access_range R2,
  class Comp = ranges::less, class Proj = identity>
  requires sortable<iterator_t<R>, Comp, Proj>
  constexpr safe_iterator_t<R>
  nth_element(R&& r, R2&& nths, Comp comp = {}, Proj proj = {});
}
\end{codeblockAdd}

@@

A: I suggest to reuse \tcode{nth_element} for discoverability but could as well be a separate name. The name \tcode{std::nth_element} is established since pre-standard STL times ($\sim$1994?). In the literature\cite{Alsuwaiyel2001,Kaligosi2006,Panh2002,lent1996,Shen1997}, it is known as \emph{select(ion)}, and the proposed algorithm is known as \emph{multiselect(ion)}.

@@

\tcode{std::multi_nth_element}

@@

\bibitem[p2375RefImpl]{p2375RefImpl}
Performance study and usage examples on \emph{this} proposal.\\
\url{https://github.com/jmlundberg/nth_element_material}

\bibitem[p2375src]{p2375src}
Document source for \emph{this} proposal including reference implementation\\
\url{https://github.com/jmlundberg/p2375}

>>>>

\begin{document}
\vspace*{-4em}
\title{\tcode{Generalisation of nth_element to a range of nths}}
\author{
Johan Lundberg\\
}
\date{} %unused. Type date explicitly below.
\maketitle
\vspace*{-3em}

\begin{tabular}{ll}
Document \#:& D2375R1\\
Date: &2021-12-28 \\
Project: & Programming Language C++ \\
Audience: & LEWG \\
Email: &\href{mailto:lundberj@gmail.com}{lundberj@gmail.com}
\end{tabular}

@@

\newpage\subsection*{Revision history vs R0}
\addcontentsline{toc}{section}{Revision history}

2021-12-16: Added clarifications and more background, explanations, wording clarifications, references, summarized performance study, applications, more questions/answers. --Based on feedback and material sent to SG9 list 2021-12-09. Added images for equalization example. Added example of \emph{Before} with projections and more examples. -- 
2021-12-26: Added explanatory images to the introduction. --  
2021-12-28: Rebased wording and synopsis to C++23 draft. Misc polishing.

@@

The use and analysis of such algorithms is widespread and mature\cite{Alsuwaiyel2001,Panh2002,lent1996,Shen1997} under the name \emph{multiple selection} or \emph{multiselect}. It is available to Python programmers as \tcode{numpy.partition}\cite{NpPart,NPImpl} since 2014.

@@

A possible implementation of the range-of-nths algorithm is provided (section \ref{Implementation}, and \cite{p2375RefImpl}).
It translates naturally to \tcode{std::ranges} versions.

@@

To clarify what is new, the addition is here called \tcode{std::multi_nth_element}, but the proposed wording overloads \tcode{std::nth_element}.

@@

As example data, consider the permuted integers [100,126) at index [0,26) and the sorted counterpart:

\includegraphics[width=0.999\textwidth]{plotting/figs/rnd.png}

\includegraphics[width=0.999\textwidth]{plotting/figs/sort.png}

@@

What the current-standard \tcode{std::nth_element} does is to rearrange the data in relation to a specified nth position, as described in the previous section. With our concrete example, with {nth = begin+7}, the effect is that the element at nth is the element (in our case: 107) that  would  be  in  that position if the whole range were sorted, and all subsequent values are no less than that value:

@@

This proposal adds support for multiple selection. That is, the possibility to provide a range of nths, such as \tcode{\{begin+7, begin+20\}} :

@@

or at \tcode{\{begin+7, begin+12, begin+20\}}:

@@

or  at \tcode{\{begin+5, begin+6, begin+14, begin+15\}}:

@@

The proposal can be implemented with an algorithm that partitions the data on the midpoint of nths using the single-nth \tcode{nth_element} and proceeds to recurse into the two partitions. Ref \cite{Alsuwaiyel2001} considers this algorithm and concludes
 complexity \mbox{O(N log m)} where \mbox{N = last - first} and m is the number of unique elements in nths.
The same paper presents a parallel version, building on refs \cite{Akl1984,Akl1989,Shen1997}. Further references are \cite{Kaligosi2006,Panh2002}.

@@

A possible implementation\footnote{
The implementation is not the point of the proposal but it may help explain it. Feedback would be very much appreciated. 
The algorithm described in \cite{Alsuwaiyel2001} is a little bit simpler since it takes \emph{unique} nths, something that is not reasonable to require here. The algorithms is \mbox{O(N log m)}. The first if-statement is found in existing single-nth \tcode{nth_element} implementations, but here it also prevents unnecessary bisections of equal and near equal nths.}\cite{p2375RefImpl}
is shown below. Comparisons and projections can also be fed through in the most natural way.

\begin{codeblock}
template< class RandomAccessIterator, class RandomAccessIterator2 >
constexpr void @\textbfx{multi_nth_element}@(RandomAccessIterator first, 
  RandomAccessIterator2 nth_first, RandomAccessIterator2 nth_last, 
  RandomAccessIterator last)
{
  if (last - first <= 32) { @\textbfx{std::sort(first, last)}@; return; }
  const auto nth_dist = nth_last - nth_first;
  if (nth_dist == 0 || *nth_first == last) return;
  const auto nth_mid = nth_first + nth_dist / 2;
  const auto at_nth_mid = *nth_mid;
  @\textbfx{nth_element(first, at_nth_mid, last)}@;
  @\textbfx{multi_nth_element(first, nth_first, nth_mid, at_nth_mid)}@;
  if (at_nth_mid != last){
    const auto nth_left = std::upper_bound(nth_mid, nth_last, at_nth_mid );
    @\textbfx{multi_nth_element(at_nth_mid + 1, nth_left, nth_last, last)}@;
  }
}

@@

\hspace{3ex}\begin{tabular}{|l|} 
  \hline 
  \textbf{Before} \\
  \hline 
\begin{codeblock} 
nth_element(v.begin(), nths[5/2-1], v.end(), pred);
nth_element(v.begin(), nths[(5/2)/2-1], nths[5/2-1], pred);
nth_element(nths[(5/2)/2-1]+1, nths[5/2-1], v.end(), pred);
nth_element(nths[(5/2)-1]+1, nths[(5/2)/2+5/2-1], nths[5-1]), pred);
nth_element(nths[(5/2)/2+5/2-1]+1, nths[5/2-1], v.end(), pred);
\end{codeblock} 
\\
\hline 
\end{tabular}

@@

\begin{codeblock}
ranges::multi_nth_element(v, vector{a,b});
processPage(a,b); // May also now continue and sort the small subrange with std::sort(a,b);
\end{codeblock}

@@

\newpage
\section{Wording and Synopsis }
\label{synopsis}

\subsection{\protect{\tcode{[alg.nth.element]}}}

\added{Underlined green text} marks additions with respect to C++23 draft (2021-12-27).

\renewcommand{\tref}[1]{ ref:\emph{#1}}
\renewcommand{\iref}[1]{ ref:\emph{#1}}

\begin{itemdescr}
\pnum
Let \tcode{comp} be \tcode{less\{\}}
and \tcode{proj} be \tcode{identity\{\}}
for the overloads with no parameters by those names.

\pnum
\expects
\range{first}{nth} and \range{nth}{last} are valid ranges.
For the overloads in namespace \tcode{std},
\tcode{RandomAccessIterator} meets
the \oldconcept{ValueSwappable} requirements\iref{swappable.requirements}, and
the type of \tcode{*first} meets
the \oldconcept{MoveConstructible} (\tref{cpp17.moveconstructible}) and
\oldconcept{MoveAssignable} (\tref{cpp17.moveassignable}) requirements.
\added{For the overloads taking a range [nths_first,nths_last), RandomAccessIterator2 is a \oldconcept{RandomAccess} iterator, and *nths_first is convertible to  RandomAccessIterator. For every iterator i and j in the range [nths_first,nths_last), it holds that if \tcodeadd{i<j} then \tcodeadd{!(*j<*i).}}

\pnum
\effects
After \tcode{nth_element} the element in the position pointed to by \tcode{nth}
is the element that would be in that position
if the whole range were sorted with respect to \tcode{comp} and \tcode{proj},
unless \tcode{nth == last}.
Also for every iterator \tcode{i} in the range \range{first}{nth}
and every iterator \tcode{j} in the range \range{nth}{last}
it holds that:
\tcode{bool(invoke(comp, invoke(proj, *j), invoke(proj, *i)))} is \tcode{false}.
\added{For the overloads taking a range of \tcodeadd{nths}, this holds for all \tcodeadd{nth} in \tcodeadd{nths}.}

\pnum
\returns
\tcode{last} for the overload in namespace \tcode{ranges}.

\pnum
\complexity
For the overloads with no \tcode{ExecutionPolicy}, linear on average.
For the overloads with an \tcode{ExecutionPolicy}, \bigoh{N} applications of
the predicate, and \bigoh{N \log N} swaps, where $N = \tcode{last - first}$.
\added{For overloads taking a range of \tcodeadd{nths} but no ExecutionPolicy the complexity is approximately \bigoh{N~log~m} where $m$ is the number of unique elements in \tcodeadd{nths}.
For overloads taking a range of \tcodeadd{nths} and an ExecutionPolicy the complexity is approximately \bigoh{N~log~N}.}
\end{itemdescr}

@@

Added signatures to \tcode{std::}

@@

template<class RandomAccessIterator, class RandomAccessIterator2, class Compare>
constexpr void nth_element(
RandomAccessIterator first, 
RandomAccessIterator2 nths_first, RandomAccessIterator2 nths_last,
RandomAccessIterator last, Compare comp);
\end{codeblockAdd}
\newpage
\begin{codeblockAdd}
template<class ExecutionPolicy, class RandomAccessIterator, class RandomAccessIterator2>
void nth_element(ExecutionPolicy&& exec,
RandomAccessIterator first, 
RandomAccessIterator2 nths_first, RandomAccessIterator2 nths_last,
RandomAccessIterator last);

@@

Added signatures to \tcode{std::ranges::}

\begin{codeblockAdd}
namespace ranges {
  template<@\libconcept{random_access_iterator}@ I, @\libconcept{sentinel_for}@<I> S,
  random_access_range R2, class Comp = ranges::less, class Proj = identity>
  requires @\libconcept{sortable}@<I, Comp, Proj>
  && std::convertible_to<decltype(*std::declval<R2>().begin()), I>
  constexpr I nth_element(I first, I nth, S last, Comp comp = {}, Proj proj = {});

  template<@\libconcept{random_access_range}@ R,
  random_access_range R2, class Comp = ranges::less, class Proj = identity>
  requires @\libconcept{sortable}@<iterator_t<R>, Comp, Proj>
  && std::convertible_to<decltype(*std::declval<R2>().begin()), 
   decltype(std::declval<R>().begin())> 
  constexpr borrowed_iterator_t<R> 
  nth_element(R&& r, iterator_t<R> nth, Comp comp = {}, Proj proj = {}); 
}

\end{codeblockAdd}
% for reference, this is the standard draft as of 2021-12-27:
%  template<@\libconcept{random_access_iterator}@ I, @\libconcept{sentinel_for}@<I> S,
%  class Comp = ranges::less, class Proj = identity>
%  requires @\libconcept{sortable}@<I, Comp, Proj>
%  constexpr I nth_element(I first, I nth, S last, Comp comp = {}, Proj proj = {});
%
%  template<@\libconcept{random_access_range}@ R, class Comp = ranges::less, class Proj = identity>
%  requires @\libconcept{sortable}@<iterator_t<R>, Comp, Proj>
%  constexpr borrowed_iterator_t<R> nth_element(R&& r, iterator_t<R> nth, Comp comp = {}, Proj proj = {});

@@

A: I suggest to reuse \tcode{nth_element} for discoverability but it could as well be a separate name. The name \tcode{std::nth_element} is established since pre-standard STL times ($\sim$1994?). In the literature\cite{Alsuwaiyel2001,Kaligosi2006,Panh2002,lent1996,Shen1997}, it is known as \emph{select(ion)}, and the proposed algorithm is known as \emph{multiselect(ion)}.

@@

@@

\bibitem[p2375RefImpl]{p2375RefImpl}
 Reference Implementation, Performance study and usage examples on \emph{this} proposal.\\
\url{https://github.com/jmlundberg/nth_element_material}

\bibitem[p2375src]{p2375src}
Document source and status page for \emph{this} proposal.\\
\url{https://github.com/jmlundberg/p2375}
